---
title: "HMD Analysis To Calculate Premature Mortality Globally"
author: "Mathew Kiang"
date: "2/15/2018"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We want to be able to look at the county SMRs, convert them to age-standardized premature mortality rates, and compare the rates of counties with that of other countries. This file just takes the Human Mortality Database files and calculates the age-standardized premature mortality rate for a bunch of different counties over the same time period. Because not every county has 2010-2015 data (i.e., some countries end in 2013 or 2014), we will calculate whatever rate we have for that period and then take the "average" annual rate. 

## Set up

All countries have their own files. All files have the same structure.
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

## Helper functions
read_hmd_file <- function(f) {
    temp_df <- read_table2(f, skip = 4, 
                           col_names = c("year", "age", "female", 
                                         "male", "total"), 
                           col_types = "ccnnn", na = c("NA", "."))
    temp_df$country <- str_split(f, "\\.|//")[[1]][[3]]
    
    return(temp_df)
}

## Get a list of death and population files
death_files <- list.files(paste0("./data_raw/", 
                                 "human_mortality_database/", 
                                 "Deaths_5x1/"), 
                          full.names = TRUE)

pop_files <- list.files(paste0("./data_raw/", 
                               "human_mortality_database/", 
                               "Population5/"), 
                        full.names = TRUE)
```

## Deaths
Note that for some years, the boundaries changed so there are duplicate counts of before (`-`) and after (`+`). I just remove the `+`/`-` and take the average. We also drop the female and male columns and just use total.

```{r}
deaths_df <- bind_rows(
    map(death_files, ~read_hmd_file(.x))
) %>% 
    rename(deaths = total) %>% 
    mutate(year = as.integer(gsub("\\+|\\-", "", year))) %>% 
    group_by(year, age, country) %>% 
    mutate_all(mean) %>% 
    ungroup() %>% 
    select(-female, -male)

print(deaths_df, n = 10)
```

## Population
Again, some years had boundary changes with `-` indicating before and `+` indicating after the boundary change for that year. We import year as a character, remove the symbol, and take the average for that year. 


```{r}
pops_df <- bind_rows(
    map(pop_files, ~read_hmd_file(.x))
) %>% 
    rename(population = total) %>% 
    mutate(year = as.integer(gsub("\\+|\\-", "", year))) %>% 
    group_by(year, age, country) %>% 
    mutate_all(mean) %>% 
    ungroup() %>% 
    select(-female, -male)

print(pops_df, n = 10)
```

## Combine them
```{r}
hmd_data <- pops_df %>% 
    left_join(deaths_df) %>% 
    select(year, age, country, deaths, population)

print(hmd_data, n = 10)
```

## Recode age to match the standard pops
```{r}
std_pop <- narcan::std_pops %>% 
    filter(standard == "s204") %>% 
    select(age, age_cat, pop_std)

print(std_pop)
```

So we need to collapse the `0` and `1-4` groups of the HMD data into one group and all the `85+` groups into one group.

```{r}
## First, make a new column with the first number of the age category
hmd_data <- hmd_data %>% 
    mutate(age_orig = age, 
           age = as.integer(
               unlist(
                   lapply(
                       str_split(age_orig, 
                                 pattern = "\\-|\\+"), 
                       function(x) x[[1]])
                   )
               )
           )

head(hmd_data)
```

```{r}
## Now create a new collapsed age column
hmd_data <- hmd_data %>% 
    mutate(new_age = case_when(
        age <= 1 ~ 0L, 
        age >= 85 ~ 85L, 
        TRUE ~ age
    ))

print(hmd_data, n = 25)
```

```{r, warning=FALSE, message=FALSE}
## Now collapse according to new age column
hmd_data <- hmd_data %>% 
    group_by(country, year, new_age) %>% 
    summarize(population = sum(population), 
              deaths = sum(deaths)) %>% 
    rename(age = new_age) %>% 
    left_join(std_pop, by = "age") %>% 
    ungroup()

head(hmd_data)
```

## Calculate average premature mortality between 2010-2015

Because not every country has all observations between 2010 and 2015, we'll just calculate premature mortality for this period for all years that are observed and then take the average. 

```{r}
sub_df <- hmd_data %>% 
    filter(year >= 2010, 
           year <= 2015, 
           age < 65) 

age_specific_rates <- sub_df %>% 
    narcan::calc_asrate_var(prem_death, 
                            death_col = deaths, 
                            pop_col = population)

head(age_specific_rates)
```

```{r}
age_std_rates <- age_specific_rates %>% 
    group_by(country, year) %>% 
    summarize(prem_std_rate = weighted.mean(prem_death_rate, pop_std)) %>% 
    ungroup()
```

```{r}
average_premature_mortality <- age_std_rates %>% 
    group_by(country) %>% 
    summarize(avg_prem = mean(prem_std_rate, na.rm = TRUE), 
              n_years = n())
```

## Add useful names
To get more useful names, we will cycle through the death files and extract the country name which should be on the first line.

```{r}
country_name <- NULL
country_abb <- NULL

for (f in death_files) {
    cname <- strsplit(
        read_lines(f, n_max = 1), 
        split = ",")[[1]][[1]]
    
    cabb <-  str_split(f, "\\.|//")[[1]][[3]]
    
    country_name <- c(country_name, cname)
    country_abb  <- c(country_abb, cabb)
}

c_map <- tibble(country_name, country_abb)
```

```{r}
average_premature_mortality <- average_premature_mortality %>% 
    left_join(c_map, by = c("country" = "country_abb")) %>% 
    arrange(avg_prem)

knitr::kable(average_premature_mortality %>% 
                 select(country_name, n_years, avg_prem), 
             digits = 1, 
             col.names = c( "Country", "Years (N)", "Average Premature Mortality Rate"))
```

## Save
```{r}
write.csv(average_premature_mortality, 
          "./data_working/hmd_countries_average_prem.csv", 
          row.names = FALSE)
```

